---
title: "CHR"
format: html
editor: visual
---

## 🌐 🌽 County Health Rankings w/ Bayesian Regression

### Libraries

```{r}
library(tidyverse)
library(janitor)
library(rstanarm)
library(here)
library(readxl)
library(maps)
library(viridis)
library(sf)
```

### Data

```{r}
chr_data <- read_excel(here("data/2025_County_Health_Rankings_Data.xlsx"), sheet = 2, skip = 1)
```

```{r}
chr_data <- chr_data %>% 
  clean_names
```

#### Regression Variables of Interest

-   `food_environment_index`

    -   The **Food Environment Index (FEI)** is a **County Health Rankings** composite (0–10 scale, where **10 = best**) that summarizes how easy it is for people in a county to get enough **healthy, affordable food**.

        It combines two pieces of information:

        1.  **Food insecurity** – the estimated share of residents who lack reliable access to enough food.

        2.  **Limited access to healthy foods** – the share of people who are **low-income and far from a grocery store** (typically ≥1 mile in urban areas or ≥10 miles in rural areas).

            These components are blended and rescaled to 0–10:

            **Higher FEI (closer to 10)** → lower food insecurity and better physical access to healthy food.

            **Lower FEI (closer to 0)** → more food insecurity and poorer access (more “food deserts”).

-   `percent_fair_or_poor_health`

    -   a County Health Rankings (CHR) indicator from the BRFSS survey:

        -   **What it is:** the **percentage of adults** who answer **“fair” or “poor”** to the question, “In general, would you say your health is excellent, very good, good, fair, or poor?”

Simplify dataframe to select columns of interest

```{r}
chr_data_simplified <- chr_data %>% 
  select(state, county, food_environment_index, percent_fair_or_poor_health)
```

### Simple scatter plot, linear model

```{r}
ggplot(data = chr_data_simplified, aes(x = food_environment_index, y = percent_fair_or_poor_health)) +
  geom_point(alpha = 0.1) +
  geom_smooth()
```

```{r}
lm(percent_fair_or_poor_health ~ food_environment_index, chr_data_simplified)
```

### Priors

Literature-Informed Priors: Fair/Poor Health \~ Food Environment Index

Empirical Background

Several studies using BRFSS, County Health Rankings (CHR), or linked datasets find that better local food access is associated with lower rates of poor self-rated health.

| Parameter | Prior (percent scale) | What it encodes | Literature tie-in |
|----|----|----|----|
| **Intercept (β₀)** | `Normal(17, 4)` | At **average FEI**, a typical county has \~**17%** fair/poor health (95% prior ≈ **9–25%**) | BRFSS/CHR descriptive range (≈14–20% typical); compatible with Berkowitz (food insecurity ↔ higher % fair/poor) |
| **Slope (β₁)** | `Normal(-1.3, 0.6)` *(pp per +1 FEI)* | Each **+1 FEI** is expected to **reduce** fair/poor health by **\~1.3 percentage points** on average (95% prior ≈ **−2.5 to −0.1** pp) | **Fan & Jin (2022)** report −**1.3 pp** per +1 FEI; **Liese et al. (2014)** (odds ≈1.1× worse per unit) implies ≈ **−1 to −1.6 pp** per +1 FEI near 15–20% baseline |
| **Residual SD (σ)** | `Exponential(1/4.5)` | Mean residual scatter ≈ **4.5 percentage points** (allows substantial county heterogeneity) | County-level BRFSS/CHR models often show **4–6 pp** residual SD after single-predictor adjustment |

### Simulation with rstanarm

```{r}
health_model <- stan_glm(percent_fair_or_poor_health ~ food_environment_index, data = chr_data_simplified,
                       family = gaussian,
                       prior_intercept = normal(17, 4),
                       prior = normal(-1.3, 0.6), 
                       prior_aux = exponential(1/4.5),
                       chains = 4, iter = 4000, seed = 84735)
```

```{r}
health_model
```

→ Each **+1** in **Food Environment Index** is associated with **\~2.6 percentage-point lower** fair/poor health, on average.

**Residual SD (sigma)** = **3.6 pp**: after accounting for FEI, counties typically vary by about ±3–4 percentage points around the line.

### Challenge

Make a map describing the **Food Environment Index** in the US

```{r}
# Create a US map showing Food Environment Index by county
# Get county map data
county_map <- map_data("county")

# Prepare the CHR data for mapping
# First, let's check what columns we have
glimpse(chr_data_simplified)

# Create mapping data by standardizing county names
mapping_data <- chr_data_simplified %>%
  filter(!is.na(food_environment_index)) %>%
  mutate(
    region = tolower(state),
    subregion = tolower(county),
    # Remove common suffixes that might not match map data
    subregion = str_remove_all(subregion, " county$| parish$| borough$| census area$| municipality$")
  ) %>%
  select(region, subregion, food_environment_index)

# Join with map data
map_plot <- county_map %>%
  left_join(mapping_data, by = c("region", "subregion"))

# Create the choropleth map
ggplot(map_plot, aes(x = long, y = lat, group = group, fill = food_environment_index)) +
  geom_polygon(color = "white", size = 0.05) +
  scale_fill_viridis_c(
    name = "Food Environment\nIndex (0-10)",
    option = "plasma",
    direction = 1,
    na.value = "grey90",
    limits = c(0, 10)
  ) +
  coord_map() +
  theme_void() +
  labs(
    title = "Food Environment Index by County",
    subtitle = "Higher values (darker colors) indicate better food access and lower food insecurity",
    caption = "Data: County Health Rankings 2025"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 12),
    plot.caption = element_text(hjust = 0.5, size = 10),
    legend.position = "bottom",
    legend.key.width = unit(2, "cm"),
    legend.title = element_text(size = 10)
  )
```

**Interpretation:** This choropleth map visualizes the Food Environment Index across US counties using real County Health Rankings data. The color gradient shows geographic patterns in food access:

-   **Darker colors (higher FEI values)**: Counties with better food access, lower food insecurity, and more grocery stores accessible to low-income residents
-   **Lighter colors (lower FEI values)**: Counties with food deserts, higher food insecurity, and limited access to healthy foods
-   **Gray areas**: Counties where data was not available or didn't match the mapping dataset

The map reveals important geographic disparities in food access, with rural areas often showing lower Food Environment Index values compared to urban and suburban counties. This spatial pattern helps explain why food environment is such an important social determinant of health across different regions of the United States.
